Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-05 MS_revised.tex
ref 47

Solanaceae plants contain chemical compounds and it has been suggested
that the high diversity of Ithomiini is consistent with the
``escape-and-radiate scenario'' due to a shift onto Solanaceae {[}12{]}
and radiation scenarios among chemically different lineages of
Solanaceae plants {[}8,44{]}. According to this theory, the shift from
Apocynaceae to Solanaceae allowed Ithomiini to invade newly available
resources due to a possible key innovation that allowed them cope with
secondary metabolites of the new hosts. Additional studies are needed to
identify the actual enzymes that Ithomiini species might be using for
detoxification of ingested food as they have been found in other
butterfly groups {[}47{]}.

@@

The diffuse cospeciation hypothesis predicts almost identical ages of
insects and their hostplants, while the ``resource abundance-dependent
diversity'' and the ``escape-and-radiate'' hypotheses state that insects
diversify after their hostplants {[}2--4{]}. Wheat et al. (2007)
{[}47{]} found strong evidence for a model of speciation congruent with
Ehrlich and Raven's hypothesis in Pieridae butterflies due to, in
addition to the identification of a key innovation, a burst of
diversification in glucosinolate-feeding taxa shortly afterwards (with a
lag of circa 10 MY). According to a recent dated phylogeny of
the Angiosperms {[}50{]}, the family Solanaceae split from its sister
group about 59 (49--68) MYA and diversification started (crown group
age) around 37 (29--47) MYA. Wahlberg et al. (2009) {[}13{]} give the
ages for origin and diversification for Ithomiini at 45 (39--53) and 37
(32--43) MYA, respectively. Thus, current evidence shows that Solanaceae
and Ithomiini might have diversified around the same time, during the
Late Eocene and Oligocene, and this would be congruent with the diffuse
cospeciation hypothesis.

>>>>

Solanaceae plants contain chemical compounds and it has been suggested
that the high diversity of Ithomiini is consistent with the
``escape-and-radiate scenario'' due to a shift onto Solanaceae {[}12{]}
and radiation scenarios among chemically different lineages of
Solanaceae plants {[}8,44{]}. According to this theory, the shift from
Apocynaceae to Solanaceae allowed Ithomiini to invade newly available
resources due to a possible key innovation that allowed them cope with
secondary metabolites of the new hosts. Additional studies are needed to
identify the actual enzymes that Ithomiini species might be using for
detoxification of ingested food as they have been found in other
butterfly groups {[}48{]}.

@@

The diffuse cospeciation hypothesis predicts almost identical ages of
insects and their hostplants, while the ``resource abundance-dependent
diversity'' and the ``escape-and-radiate'' hypotheses state that insects
diversify after their hostplants {[}2--4{]}. Wheat et al. (2007)
{[}48{]} found strong evidence for a model of speciation congruent with
Ehrlich and Raven's hypothesis in Pieridae butterflies due to, in
addition to the identification of a key innovation, a burst of
diversification in glucosinolate-feeding taxa shortly afterwards (with a
lag of circa 10 MY). According to a recent dated phylogeny of
the Angiosperms {[}50{]}, the family Solanaceae split from its sister
group about 59 (49--68) MYA and diversification started (crown group
age) around 37 (29--47) MYA. Wahlberg et al. (2009) {[}13{]} give the
ages for origin and diversification for Ithomiini at 45 (39--53) and 37
(32--43) MYA, respectively. Thus, current evidence shows that Solanaceae
and Ithomiini might have diversified around the same time, during the
Late Eocene and Oligocene, and this would be congruent with the diffuse
cospeciation hypothesis.
